Galaxy image statistical generator: how to be confident?

### 2024-09-19 paper/main.tex
more results

The ability of generative models have been rapidly adopted in many domains. For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}. Concerning Astrophysics and Cosmology the generative models are used in different tasks as deblending \citep{Hemmati_2022,Arcelin2020} {\color{red} à compléter}) and what we are concern in this article the synthetic galaxy image generations  going beyond parametrized analytic light profile simulations (e.g. \texttt{GalSim} by \cite{ROWE2015121}). Along this line \citep{ravanbakhsh2016,Fussell2019} use GAN models, \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model.

@@

\cite{kadkhodaie2024generalization} address more mathematically oriented questions and demonstrate the passage between memorization to generalization of  diffusion generative models when the dataset size increases and propose interpretation of what is learned by the network as a sort of \textit{geometry-adaptive harmonic basis} going beyond the commonly used \textit{steerable wavelet basis} first introduced and studied in \cite{Simoncelli1995b,Unser2013} and developped for instance in the \texttt{Kyamato} library \citep{JMLR:v21:19-047}. The authors uses \textit{denoiser} architectures as \texttt{UNet} \citep{2015arXiv150504597R} and \texttt{BF-CNN} networks \citep{Mohan2020Robust} trained on \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} bedroom  \citep{Yu2015} reduced datasets of $O(10^5)$ images resized to $80\times 80$ pixels (grayscale). The dataset size and the number of pisels per image make possible to investigate how the \texttt{Glow} normalizing flow based model compare to denoiser based diffusion model concerning the generation of galaxy images both trained with the same dataset issued from DR7 SDSS survey.

In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some guide lines. Appendix gives more details on the dataset used.

@@

It is usual to call $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part. The different VAE models differ by the distribution families used. The simplest choice may be commonly-used Gaussian multivariate distributions, and the centred isotropic with unit variance version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb. $\bm{\theta}_1$ are missing is this scenario). Once the optimization is done, the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples. It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of \textit{posterior collapse} which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}. 
%
\subsection{Generative Adversarial Networks}
%
The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$. To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution. To reach that goal one solves the \texttt{min-max} problem
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\}
\end{equation} 
where $p_{data}$ is the data generating distribution. Both the  $\bm{G}$ and $\bm{D}$ are parametrized networks. Notice that at parameter of $\bm{G}$ fixed, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})}
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}} |^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right) }  + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\}
\end{equation}
leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution. 
%
\subsection{Normalizing Flows}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached in case of $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))=0$ which is true iif $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. But this situation cannot be true in general as typical $p_{\bm{\phi}}$ parametrizations involve for instance independent Gaussian distributions. The limited choices of the parametrization is one of the limitation of VAE. So, one motivates the use of normalizing flows (NF) to allow for more flexibility in the distribution family design to hope that the true posterior is member of such family \citep{Tabak2010, Tabak2013a, Rezende2015}.

Independently of the VAE context, a flow $T$ is defined as diffeomorphism (aka bijector) between the data space and a latent space such that
\begin{align}
\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x})
\end{align}
So, if $\bm{z}$ distribution is $\pi(\bm{z})$ (e.g. $\mathcal{N}(\bm{z},\bm{0},\bm{1})$) then the following relation apply
\begin{equation}
p(\bm{x}) =  \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|
\end{equation}
Contrary to VAE and GAN the latent and data space dimensionalities are the same. The flow $T$ is in general a composition of several individual flows $\{T_i\}_{i<n}$ in such a way that
\begin{align}
T=T_1\circ T_2\circ \dots \circ T_n & \quad \Leftrightarrow \quad 
T^{-1}=T_n^{-1}\circ T_{n-1}^{-1}\circ \dots \circ T_1^{-1}
\end{align}
Figure \ref{fig-normflow} represents a schematic views of the \textit{forward} or \textit{generative} direction and the \textit{backward} or \textit{training} direction. 
If one notes $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i<n$  (using $\bm{z}_0=\bm{z}$, $\bm{z}_n=\bm{x}$) then
\begin{equation}
\log |\mathrm{det} J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|
\label{eq-flow-jacob}
\end{equation}
Theorem exists showing that under moderate assumptions flow-based models are able to represent any density distributions \citep{Bogachev2005,huang2019solving}. So, if the flows are well chosen, one can sample $\bm{x}$ with potentially complex (multi-modes) density distribution from a simple spherical centred multivariate Gaussian distribution.

@@

An answer has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes.  
%
\subsection{Datasets and Models}
%
We propose to replay their experiment using the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. For comparison purposes, we also use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019} for convenience. The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.

@@

We are more concerned with galaxy image generation, but as we need to train two models with up to $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} but first all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels and from the $(z,g,r)$ channels we produce a single  channel in gray levels in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray level is coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}. With such dataset according to \cite{kadkhodaie2024generalization} results, the transition between memorization to generalisation regimes is expected to occur with a number of training images in the range $10^4-10^5$.

So in the same spirit as \citep{kadkhodaie2024generalization}, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images. The same training procedure used by \cite{kadkhodaie2024generalization} has been applied. Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations. {\color{red} We refer the reader to the  github repository companion for the details of the different codes and notebooks used and models trained as well as the training dataset of $250,000$ galaxies.}
%
\subsection{Results}
%
\subsubsection{Sample images}
Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and from a diffusion-based and flow-based models generation using $N=10^5$ trained images. 
%
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-original_samples_gray.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_gray.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
        \centering
        \raisebox{0mm}{\includegraphics[width=0.8\textwidth]{fig-glow_samples_gray.pdf}}
        %\caption{xxx}
    \end{subfigure}
    \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images. On the left some original images are also displayed for visual comparison.}
    \label{fig-Original-Glow-UNet-samples}
\end{figure}
At first look the generated images may be not distinguishable from the original images. Notice that we do not expect high resolution images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality. Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing?
%
\subsubsection{U-Net denoising performances}
Figures \ref{fig-UNet-denoising-train},\ref{fig-UNet-denoising-test}...
\begin{figure}
    \centering
		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_train.pdf}\\
		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100_denoising_train.pdf}
	\caption{Performances of a some \texttt{U-Net} models using an image of the \textit{training sample}. Top row: (left to right) the original image with progressively a larger noise added. \texttt{PSNR} is defined as $10 \log10$ ratio of squared dynamic range
to mean square error; the distance is the euclidean distance divided between the square root of number of pixels. Middle row: (left to right) (image, PSNR, distance) obtained after denoising the corresponding image of top row with \texttt{U-Net} trained with  $N=100$ images. Bottom row: same as the middle row but   with \texttt{U-Net} trained with  $N=100,000$ images}
	\label{fig-UNet-denoising-train}
\end{figure}

\begin{figure}
    \centering
		\includegraphics[width=0.8\textwidth]{fig-UNet_100000_denoising_test.pdf}\\
		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_test.pdf}
	\caption{Similar to  Figure \ref{fig-UNet-denoising-train} but  using an image never used to train any models.}
	\label{fig-UNet-denoising-test}
\end{figure}
%
\subsubsection{Cosine similarity metric}
Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom). Despite the fact that the original training images are less diverse as e.g. the \texttt{CelebA} face or the \texttt{LSUN} bedroom images, we experience the following facts: 
\begin{itemize}
\item first: at low sized dataset ($N=100$), the samples generated by the two models look very similar to a training image (cosine similarity close to 1) as shown by bottom left histogram while the sample generated from the same white noise image looks different as the cosine similarity is different from 1 (upper left histogram);
\item second: greater the dataset size is then more and more generated images looks different (from left to right histograms of the lower row) and simultaneously the samples of the two models are more and more similar (from left to right histograms of the upper row). Already for $N=100,000$ we are close to get a unique bin filled at cosine similarity of 1. 
\end{itemize}

\begin{figure}
    \centering
	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-AB.pdf}\\
	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-ATrain.pdf}
	\caption{Evolution of the cosine similarity metric (see text) considering two diffusion-based models trained with two independent datasets of size $N$. On the upper plots are displayed normalized histograms considering $1,000$ generated samples for each of the two models \textit{using as seed the same white noise image}.  On the lower plots, this is the maximum of the "cosine" metric which is considered using the 1,000 generated samples of one model compared to all images of its corresponding training dataset (in other words this is the "cosine" metric of the closest training image from the generated sample). The logarithmic vertical and linear horizontal (cosine value) ranges of each raw are shared among the different histograms. Notice that the horizontal range of the bottom row is restricted to a range closer to the maximum value (i.e. 1) materialized by the red vertical line on each histogram. This figure has been inspired by Figure 2 of \cite{kadkhodaie2024generalization}.}
	\label{fig-cosine-diffusion}
\end{figure}
%
The first fact tell us that clearly the generative model are simply reproducing the dataset (so-called \textit{memorization} regime), while the second fact is an indication of a rather smooth transition towards a \textit{generation} regime when the dataset is increasing. These results are in perfect agreement with the ones exposed in \cite{kadkhodaie2024generalization}. What is less expected is that we still need $10^5$ samples to be confident that the generation regime applies.

This last remarks invite ourselves to perform the same experiment with the \texttt{Glow} flow-based model with $N=10^5$ trained images. The result is shown on Figure \ref{fig-cosine-glow}. It is quite remarkable that even if the generated samples seems different from any of the trained images\footnote{Also any of horizontal flipped, vertical flipped or horizontal+vertical flipped versions.} as right histogram does not peak at unity cosine similarity, in the same time even seeded by the same white noise image the two models output images are different as the left histogram does not show a peak at unity cosine similarity too. This is clearly a sign that the two models haven't leaned the same forward transport mapping. 
%
\begin{figure}
    \centering
	\includegraphics[width=0.8\textwidth]{fig-glow-cosine-AB_ATrain.pdf}
	\caption{Histograms similar to those of Figure \ref{fig-cosine-diffusion} for two Glow models using two different datasets of size $N$ but the same as the ones used by the diffusion models. Left: the cosine similarity of $1,000$ samples generated by the two models seeded by the same white noise image. Right: the cosine similarity of $1,000$ samples generated by one of the models and the training images and their flipped versions (horizontal, vertical and horizontal+vertical).}
	\label{fig-cosine-glow} 
\end{figure}

>>>>

The ability of generative models have been rapidly adopted in many domains. For instance, in High Energy Physics the reader may be interested by the review \cite{PhysRevD.107.076017}. Concerning galaxy image generation several architectures have been used and some times in conjunction of deblending and deconvoling tasks. For instance \cite{ravanbakhsh2016} use conditional VAE and GAN models, \citep{Schawinski2017,Fussell2019,Hemmati_2022} use GAN models as well and \citep{Arcelin2020} use a VAE model and  \cite{Lanusse2021} propose an hybrid VAE-Normalizing Flow architecture and \cite{smith2021} use a denoising stochastic diffusion model. These generator are intended to go beyond parametrized analytic light profile simulations (e.g. \texttt{GalSim} by \cite{ROWE2015121}) to capture complex structures and interplay between background and single galaxy or blended galaxies signals.

@@

\cite{kadkhodaie2024generalization} address more mathematically oriented questions and demonstrate the passage between memorization to generalization of  diffusion generative models when the dataset size increases and propose interpretation of what is learned by the network as a sort of \textit{geometry-adaptive harmonic basis} going beyond the commonly used \textit{steerable wavelet basis} first introduced and studied in \cite{Simoncelli1995b,Unser2013} and developped for instance in the \texttt{Kyamato} library \citep{JMLR:v21:19-047}. The authors uses \textit{denoiser} architectures as \texttt{U-Net} \citep{2015arXiv150504597R} and \texttt{BF-CNN} networks \citep{Mohan2020Robust} trained on \texttt{CelebA} \citep{Liu2015} and \texttt{LSUN} bedroom  \citep{Yu2015} reduced datasets of $O(10^5)$ images resized to $80\times 80$ pixels (grayscale). The dataset size and the number of pisels per image make possible to investigate how the \texttt{Glow} normalizing flow based model compare to denoiser based diffusion model concerning the generation of galaxy images both trained with the same dataset issued from DR7 SDSS survey.

In the following sections, first we briefly describe the different kind of generative models and we show the outcome of the conducted numerical experiment before drawing some conclusion.

@@

It is usual to call $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part. The different VAE models differ by the distribution families used. The simplest choice may be commonly-used Gaussian multivariate distributions, and the centred isotropic with unit variance version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb. $\bm{\theta}_1$ are missing is this scenario). Once the optimization is done, the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples. It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of \textit{posterior collapse} which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}. 
%
\subsection{Generative Adversarial Networks}
%
The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$. To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution. To reach that goal one solves the \texttt{min-max} problem
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\}
\end{equation} 
where $p_{data}$ is the data generating distribution. Both the  $\bm{G}$ and $\bm{D}$ are parametrized networks. Notice that at parameter of $\bm{G}$ fixed, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})}
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}} |^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right) }  + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\}
\end{equation}
leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution.

GAN architectures have been used for galaxy generation as \citep[e.g.][]{Schawinski2017,Coccomini2021,Hemmati_2022} in conjunction of deblending and deconvolving tasks. 
%
\subsection{Normalizing Flows}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached in case of $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))=0$ which is true iif $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. But this situation cannot be true in general as typical $p_{\bm{\phi}}$ parametrizations involve for instance independent Gaussian distributions. The limited choices of the parametrization is one of the limitation of VAE. So, one motivates the use of normalizing flows (NF) to allow for more flexibility in the distribution family design to hope that the true posterior is member of such family \citep{Tabak2010, Tabak2013a, Rezende2015}.

A flow $T$ is defined as diffeomorphism (aka bijector) between the data space and a latent space such that
\begin{align}
\bm{x} = T(\bm{z}) & \quad \Leftrightarrow \quad \bm{z} = T^{-1}(\bm{x})
\end{align}
So, if $\bm{z}$ distribution is $\pi(\bm{z})$ (e.g. $\mathcal{N}(\bm{z},\bm{0},\bm{1})$) then the following relation apply
\begin{equation}
p(\bm{x}) =  \pi(\bm{z})|\det J_T(\bm{z})|^{-1} = \pi(T^{-1}(\bm{x}))|\det J_{T^{-1}}(\bm{x})|
\end{equation}
Contrary to VAE and GAN the latent and data space dimensionalities are the same by definition. The flow $T$ is in general a composition of several individual flows $\{T_i\}_{i<n}$ in such a way that
\begin{align}
T=T_1\circ T_2\circ \dots \circ T_n & \quad \Leftrightarrow \quad 
T^{-1}=T_n^{-1}\circ T_{n-1}^{-1}\circ \dots \circ T_1^{-1}
\end{align}
Figure \ref{fig-normflow} represents a schematic views of the \textit{forward} or \textit{generative} direction and the \textit{backward} or \textit{training} direction. 
If one notes $\bm{z}_i = T_i(\bm{z}_{i-1})$ for all $i<n$  (using $\bm{z}_0=\bm{z}$, $\bm{z}_n=\bm{x}$) then
\begin{equation}
\log |\mathrm{det} J_T| = \sum_{i=0}^{n-1} \log |\det J_{T_i}(\bm{z}_{i-1})|
\label{eq-flow-jacob}
\end{equation}
Theorem exists showing that under moderate assumptions flow-based models are able to represent any density distributions \citep{Bogachev2005,huang2019solving}. So, if the flows are well chosen, one can sample $\bm{x}$ with potentially complex (multi-modes) density distribution from a simple spherical centred multivariate Gaussian distribution.

@@

%
\subsection{Datasets and Models}
%
An answer to the previous question has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes. We propose to replay their experiment using first the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. Second, for comparison purposes, we use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019}. The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.

@@

We are more concerned with galaxy image generation, but as we need to train two models with up to $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing. First, all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels, and secondly from the $(z,g,r)$ channels we produce a single  channel in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray levels are coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}. With such dataset according to \cite{kadkhodaie2024generalization} results, we expect the transition between memorization to generalisation regimes to occur with a number of training images in the range $10^4-10^5$.

So, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images. The same training procedure used by \cite{kadkhodaie2024generalization} has been applied. Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations. {\color{red} We refer the reader to the  github repository companion of the present article for the details of the different codes and notebooks used and models trained as well as to get the whole dataset of $250,000$ original galaxy images.}
\subsection{Results}
\subsubsection{Generated images}
Figure \ref{fig-Original-Glow-UNet-samples} presents some galaxy images extracted from the original dataset and generated by  diffusion-based and flow-based models trained with $N=10^5$ original images. 
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-original_samples_gray.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
    \begin{center}
        \centering
        \includegraphics[width=0.8\textwidth]{fig-diffusion_samples_gray.pdf}
        %\caption{xxx}
    \end{center}
    \end{subfigure}%
    \begin{subfigure}[t]{0.30\columnwidth}
        \centering
        \raisebox{0mm}{\includegraphics[width=0.8\textwidth]{fig-glow_samples_gray.pdf}}
        %\caption{xxx}
    \end{subfigure}
    \caption{Illustration of the generation of some samples from the diffusion-based (\texttt{U-Net} denoiser) and the \texttt{Glow} flow-based models trained with $N=10^5$ images. On the left some original images are also displayed for visual comparison.}
    \label{fig-Original-Glow-UNet-samples}
\end{figure}
At first look the generated images may be not distinguishable from the original images.
To be a bit more quantitative, Figure \ref{fig-morpho-coeff} presents  some morphological coefficients that have been computed using \texttt{Statmorph}\footnote{\url{https://github.com/vrodgom/statmorph}} \citep{2019MNRAS.483.4140R}: the Concentration, Asymmetry and Smoothness \citep{2000AJ....119.2645B,2003ApJS..147....1C,2004AJ....128..163L}  as well as the Gini and M20 coefficients \citep{2004AJ....128..163L,10.1093/mnras/stv2078}.
\begin{figure}
    \centering
		\includegraphics[width=0.5\textwidth]{fig-CAS-all.pdf}
		\includegraphics[width=0.35\textwidth]{fig-gini-m20-all.pdf}
	\caption{Corner plots \citep{corner2016} of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-samples}). The figure has been inspired by \cite{HACKSTEIN2023100685}.}
	\label{fig-morpho-coeff}
\end{figure}
The distributions derived from the \texttt{U-Net} diffusion model are in better agreement with those obtained with the original images than are those derived from the \texttt{Glow} flow model, although this is not dramatic.
Notice that we do not expect high quality images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality. Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing with moderate dataset size?
\subsubsection{\texttt{U-Net} denoising performances}
Before diving into the generative performances of the models, we present first some results concerning the denoising performances of two \texttt{U-Net} networks trained with different dataset sizes. Figure \ref{fig-UNet-denoising-train} shows how is denoised an original image of the \textit{training set} of both networks. By contrast, to produce the Figure \ref{fig-UNet-denoising-test}, the input original image has never been used to train any networks, this is somewhat a \textit{validation image}. For both figures: the peak signal-to-noise ratio (PSNR) compared to the original image is listed above each image; the top row is composed of the input images (original or noisy) given to a denoiser; for the middle row each image is the output of the \texttt{U-Net} network trained with $N=100$ images, similarly for the bottom row the \texttt{U-Net} network has been trained with $N=100,00$ images. We remind that the networks have been trained with noisy images without any knowledge of the PSNR value (or noise variance).

Using an image from the \textit{training} dataset, both networks have similar denoising results in terms of the evolution of the output images  PSNR values with respect to the PSNR of the noisy input images. What is the remarkable by contrast is the differences of the  PSNR evolutions when considering the \textit{validation image}: if the \texttt{U-Net} network trained with $N=100,000$ images obtains similar results as for the training image which is the sign of a good generalisation power, this is  not the case for the \texttt{U-Net} network trained with $N=100$ images. For this later  network, the PSNR values of the output images are nearly constant ($\mathrm{PSNR}=27$), even for very low noise level added to the original image. This difference of denoising power between the two networks has an impact on generative process as it starts from pure noise image to progressively output a noiseless image.
\begin{figure}
    \centering
%		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_train.pdf}\\
%		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_train.pdf}
		\includegraphics[width=0.8\textwidth]{fig-UNet_denoising_train.pdf}
	\caption{Performances of a some \texttt{U-Net} models using an image of the \textit{training sample}. Top row: (left to right) the original image with progressively a larger noise added. \texttt{PSNR} is defined as $10 \log10$ ratio of squared dynamic range
to mean square error; the distance is the euclidean distance divided between the square root of number of pixels. Middle row: (left to right) (image, PSNR, distance) obtained after denoising the corresponding image of top row with \texttt{U-Net} trained with  $N=100$ images. Bottom row: same as the middle row but   with \texttt{U-Net} trained with  $N=100,000$ images.}
	\label{fig-UNet-denoising-train}
\end{figure}

\begin{figure}
    \centering
%		\includegraphics[width=0.8\textwidth]{fig-UNet_100_denoising_test.pdf}\\
%		\includegraphics[width=0.8\textwidth, trim={0 0 0 25cm},clip=True]{fig-UNet_100000_denoising_test.pdf}
	\includegraphics[width=0.8\textwidth]{fig-UNet_denoising_test.pdf}
	\caption{Similar to  Figure \ref{fig-UNet-denoising-train} but  using an image never used to train any models (i.e. a \textit{validation} image). }
	\label{fig-UNet-denoising-test}
\end{figure}
%
\subsubsection{Evolution of the generative power}
Figure \ref{fig-cosine-diffusion} explores the evolution of the cosine similarity metric \citep{books/aw/TanSK2005} considering either $1,000$ samples generated by the two diffusion-based models trained with independent dataset of same size $N$ and importantly \textit{seeded by the same white noise images} (top), or considering a single model and finding the closest training image (bottom). Despite the fact that the original training images are less diverse as e.g. the \texttt{CelebA} face or the \texttt{LSUN} bedroom images, we experience the following facts: 
\begin{itemize}
\item first: at low sized dataset ($N=100$), the samples generated by the two models look very similar to a training image (cosine similarity close to 1) as shown by bottom left histogram while the samples generated from the same white noise image looks different as the cosine similarity is different from 1 (upper left histogram);
\item second: greater the dataset size is then more and more generated images looks different (from left to right histograms of the lower row) and simultaneously the samples of the two models are more and more similar (from left to right histograms of the upper row). With $N=100,000$, we are close but not yet fully attained to the situation where the two models produce exactly the same output image (in this case a unique bin would be filled at cosine similarity of 1). 
\end{itemize}

\begin{figure}
    \centering
	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-AB.pdf}\\
	\includegraphics[width=0.8\textwidth]{fig-diffusion-cosine-ATrain.pdf}
	\caption{Evolution of the cosine similarity metric considering two diffusion-based models trained with two independent datasets of size $N$. On the upper plots are displayed normalized histograms considering $1,000$ generated samples for each of the two models \textit{using as seed the same white noise image}.  On the lower plots, this is the maximum of the metric  using the 1,000 generated samples of one model compared to all images of its corresponding training dataset (in other words this is the cosine of the closest training image from the generated sample). The logarithmic vertical and linear horizontal (cosine value) ranges of each raw are shared among the different histograms. Notice that the horizontal range of the bottom row is restricted to a range closer to the maximum value (i.e. 1) materialized by the red vertical line on each histogram. This figure has been inspired by Figure 2 of \cite{kadkhodaie2024generalization}.}
	\label{fig-cosine-diffusion}
\end{figure}
%
The first fact tell us clearly that the \texttt{U-Net}-based diffusion generative model trained with very few images are simply reproducing the dataset (so-called \textit{memorization} regime), while the second fact is an indication of a rather smooth transition towards a \textit{generation} regime when the dataset size is increasing. These results are in perfect agreement with the ones exposed in \cite{kadkhodaie2024generalization} and originate from the denoising power strength exposed in the previous section. What is surprising is that we still need $10^5$ samples to be confident that the generation regime applies with low complexity and resolution images compared to face or bedroom images.
